\section*{Exercises}
\begin{enumerate}
\item Prove that $A\bullet B$ is distributive and commutative, but non-associative.
\item Prove that $A\bullet (A\bullet A)=(A\bullet A)\bullet A$ and that 
for any natural number $n$, $A^{\bullet n}$ is well-defined (power-associative).

\item Prove that $[A,B]_+$ is distributive, $[A,A]=0$, but non-associative.

\item Prove that $//a,b,c/,d,e/=/a,\backslash d,b,c\backslash,e/=/a,b,/c,d,e//$.

\item For symmetric matrices $A$ and $B$, prove that $U_A(B)\defeq ABA$ is symmetric.
(This is the quadratic Jordan product).  Prove that $B\mapsto U_A(B)$ is linear.  What 
about $A\mapsto U_A(B)$?

\item Prove that $2\{A,B,C\}=U_{A+C}(B)-U_A(B)-U_C(B)$.


\item Consider $\mathbb{M}_{2d}(K)$ the set of $(d\times d)$-matrices with values 
in a field $K$.  Let $M^{\dagger}$ be the transpose and $J=\begin{bmatrix} 0 & I_n\\ I_n & 0 \end{bmatrix}$.
For $A\in \mathbb{M}_{2d}(K)$, define $A^*=J^{-1} A^{\dagger}J$.  Prove that each 
of the following sets is closed under addition.
\begin{align*}
    \mathbb{M}_{2d}(K)^+ & = \{ A\mid A=A^*\}\\
    \mathbb{M}_{2d}(K)^- & = \{ A\mid A=-A^*\}\\
\end{align*}
Find a distributive product which makes $\mathbb{M}_{2d}(K)^{\epsilon}$ closed.  Is either associative?

\item Let $U$, $V$, and $W$  be abelian groups and $*:U\times V\bmto W$ distributive product, i.e.\ 
\begin{gather*}
    (u+\acute{u})*v = u*v+\acute{u}*v \qquad 
    u*(v+\acute{v})= u*v+u*\acute{v}
\end{gather*}
Define 
\begin{gather*}
    \Adj(d) =\{
        (f:U\to U, g:V\to V)\mid 
        f(u)*v = u*g(v)
    \}
\end{gather*}
Show $\Adj(*)$ is closed under addition.  Then develop a binary distributive product 
for $\Adj(*)$ where $(1,1)$ behaves as the identity.  Is it associative?

\item Let $A,B$ be square complex matrices where $A^*=\bar{A}^{\dagger}=A$ and 
$B^*=B$.  Show that $(i[A,B])^*=i[A,B]$.  This is what connects Jordan algebras 
with Lie algebras in quantum mechanics.

\item Given a field $K$ and integers $d_1,\ldots,d_{\ell}$, define a 
product $(\ldots):K^{d_1}\times \cdots \times K^{d_{\ell}}\bmto K^{d_1\times \cdots\times d_{\ell}}$
where for each $1\leq a\leq \ell$,
\[
    (u_a+\acute{u}_a, u_{\bar{a}})  = (u_a,u_{\bar{a}})+(\acute{u}_a,u_{\bar{a}})
\]
Where $(u_a,u_{\bar{a}})\defeq (u_1,\ldots, u_{a}, \ldots, u_{\ell})$.
\end{enumerate}