\section{Childhood Induction}
\begin{quote}
``... small number
of symbols and their grammar are enough to capture the huge
variety of equations...''
\end{quote}

\index{evaluation}\index{induction}
The point of a variable is to replace it.  So in the formula 
$x(x+3)^2$ replacing $x$ for $7$ gives us $7(7+3)^2$.  
Yet even in the land of pure algebra where every symbol is variable it is absurd to
replace $+$ for $7$ to get $x(x73)^2$.   There is restraint in the substitution
of algebra built into what makes something an algebraic formula. When we pull 
on this thread we unravel an expansive relationship between free algebras and induction
and their role in computation.

\index{PEMDAS}\index{order of operarations}
Consider how we know what to do when   
calculating $7(7+3)^2$.  For some of us a mnemonic springs to mind
(\emph{Please Excuse My Dear Aunt Sally}) or an acronym (PEMDAS).\footnote{My 
students kindly informed me this is badge of my generation.  Now many new acronyms 
(BEMDAS) and (GEMDAS) are preferred.}
These both unwind to tell us: Parenthesis Exponents Multiplication Division Addition Subtraction
in that order. This meandering thought process somehow elucidates how to read a formula. 
It gives the symbols complex structure that 
can be visualized with the following comic strip.
\begin{center}
    \begin{tikzpicture}[yscale=0.65]
        \node (A) at (0,0) {\begin{tikzpicture}[yscale=0.75]
        \node (f) at (0,0) {$7(7+3)^2$};
        \node[below of=f,scale=0.75] {$\times$};
        \node (x1) at (-1,-2) {$7$};
        \node (sqrt1) at (1,-2) {$(7+3)^{2}$}; 
        % \node[below of=sqrt1,scale=0.75] {$\circ$};
        \node (su) at (1.5,-3) {\textasciicircum $2$};
        \node (u) at (1,-4) {$7+3$};
        \node (x2) at (0,-6) {$7$};
        \node[below of=u,scale=0.75] {$+$};
        \node (three) at (2,-6) {$3$};
        % \node (x3) at (0,-8) {$x$};
        % \node (x4) at (2,-8) {$x$};
        % \node[below of=x2,scale=0.75] {$\times$};

        \draw[-] (f) -- (x1);
        \draw[-] (f) -- (sqrt1);
        % \draw[-] (sqrt1) -- (su);
        \draw[-] (sqrt1) -- (u);
        \draw[-] (u) -- (x2);
        \draw[-] (u) -- (three);
        % \draw[-] (x2) -- (x3);
        % \draw[-] (x2) -- (x4);

    \end{tikzpicture}};
    
    \node[right of=A,xshift=3cm] (B) {\begin{tikzpicture}[yscale=0.75]
        \node (f) at (0,0) {$7(7+3)^2$};
        \node[below of=f,scale=0.75] {$\times$};
        \node (x1) at (-1,-2) {$7$};
        \node (sqrt1) at (1,-2) {$(7+3)^{2}$}; 
        % \node[below of=sqrt1,scale=0.75] {$\circ$};
        \node (su) at (1.5,-3) {\textasciicircum $2$};
        \node (u) at (1,-4) {$10$};

        \draw[-] (f) -- (x1);
        \draw[-] (f) -- (sqrt1);
        % \draw[-] (sqrt1) -- (su);
        \draw[-] (sqrt1) -- (u);

    \end{tikzpicture}};
    
    \node[right of=B, xshift=3cm] (C) {\begin{tikzpicture}[yscale=0.75]
        \node (f) at (0,0) {$7(7+3)^2$};
        \node[below of=f,scale=0.75] {$\times$};
        \node (x1) at (-1,-2) {$7$};
        \node (sqrt1) at (1,-2) {$100$}; 

        \draw[-] (f) -- (x1);
        \draw[-] (f) -- (sqrt1);

    \end{tikzpicture}};
    \node[right of=C,xshift=1cm] {$700$};

    \draw[thick] (A.north east) -- (A.south east);
    \draw[thick] (B.north east) -- (B.south east);
    \draw[thick] (C.north east) -- (C.south east);
\end{tikzpicture}
\end{center}
These step-by-step instructions start at the leaves $7$ and $3$ and join them as $7+3$ (computing $10$),
then the next step is to square (now $100$), then multiply by $7$, we reach $700$.
In hindsight, PEMDAS taught children a complicated form of induction.

