\chapter{Grammar school}

\begin{quote}
``... small number
of symbols and their grammar are enough to capture the huge
variety of equations...''
\end{quote}

The point of a variable is to replace it.  So in the formula 
$x(x+3)^2$ replacing $x$ for $7$ gives us $7(7+3)^2$.  
Yet even in the land of algebra where every symbol is variable it is absurd to
replace $+$ for $7$ to get $x(x73)^2$.   There is restraint in the substitution
of algebra built into what makes something an algebraic formula.

% It is none other than grammar.

To highlight the rules of algebra consider how we know what to do when   
calculating $7(7+3)^2$.  For some of us a mnemonic springs to mind
(\emph{Please Excuse My Dear Aunt Sally}) or an acronym (PEMDAS).
These both unwind to tell use Parenthesis Exponents Multiplication Division Addition Subtraction
in that order. This meandering thought process somehow elucidates how we read complex formulas. 
The complexity 
can be visualized with a diagram called a \emph{parse tree}.
\begin{center}
    \begin{tikzpicture}
        \node (A) at (0,0) {\begin{tikzpicture}[yscale=0.75]
        \node (f) at (0,0) {$7(7+3)^2$};
        \node[below of=f,scale=0.75] {$\times$};
        \node (x1) at (-1,-2) {$7$};
        \node (sqrt1) at (1,-2) {$(7+3)^{2}$}; 
        % \node[below of=sqrt1,scale=0.75] {$\circ$};
        \node (su) at (1.5,-3) {\textasciicircum $2$};
        \node (u) at (1,-4) {$7+3$};
        \node (x2) at (0,-6) {$7$};
        \node[below of=u,scale=0.75] {$+$};
        \node (three) at (2,-6) {$3$};
        % \node (x3) at (0,-8) {$x$};
        % \node (x4) at (2,-8) {$x$};
        % \node[below of=x2,scale=0.75] {$\times$};

        \draw[-] (f) -- (x1);
        \draw[-] (f) -- (sqrt1);
        % \draw[-] (sqrt1) -- (su);
        \draw[-] (sqrt1) -- (u);
        \draw[-] (u) -- (x2);
        \draw[-] (u) -- (three);
        % \draw[-] (x2) -- (x3);
        % \draw[-] (x2) -- (x4);

    \end{tikzpicture}};
    
    \node[right of=A,xshift=3cm] (B) {\begin{tikzpicture}[yscale=0.75]
        \node (f) at (0,0) {$7(7+3)^2$};
        \node[below of=f,scale=0.75] {$\times$};
        \node (x1) at (-1,-2) {$7$};
        \node (sqrt1) at (1,-2) {$(7+3)^{2}$}; 
        % \node[below of=sqrt1,scale=0.75] {$\circ$};
        \node (su) at (1.5,-3) {\textasciicircum $2$};
        \node (u) at (1,-4) {$10$};

        \draw[-] (f) -- (x1);
        \draw[-] (f) -- (sqrt1);
        % \draw[-] (sqrt1) -- (su);
        \draw[-] (sqrt1) -- (u);

    \end{tikzpicture}};
    
    \node[right of=B, xshift=3cm] (C) {\begin{tikzpicture}[yscale=0.75]
        \node (f) at (0,0) {$7(7+3)^2$};
        \node[below of=f,scale=0.75] {$\times$};
        \node (x1) at (-1,-2) {$7$};
        \node (sqrt1) at (1,-2) {$100$}; 

        \draw[-] (f) -- (x1);
        \draw[-] (f) -- (sqrt1);

    \end{tikzpicture}};
    \node[right of=C,xshift=1cm] {$700$};
\end{tikzpicture}
\end{center}
We can read the tree like step-by-step instructions.  Start at the leaves and join
them by whatever operation is displayed on adjacent branches.
We start at the bottom with $7,3$, and join them as $7+3$ (computing $10$),
then the next step is to square (now $100$), then multiply by $7$, we reach $700$.
In hindsight, we taught children a complicated form of induction.

You may have been taught induction through stories of falling 
dominos.  Good.  But what if induction was more like climbing, 
and the domino illustration was bottling up the experience
of climbing stairs?  Surely its more fun to climb trees and mountains.
Setting up this induction was nothing more than a fragment of text but 
read through the lens of a grammar, PEMDAS.
% Climbing can go up (induct) or down (recurse), but of course I drew the 
% tree upside-down; so, the metaphor must be rotated.
% My goal is to turn all of your views on induction on their head.
% There are many inductions, one for each free algebra, but that's the forest 
% and we should look first at some more trees.

\begin{quote}
    \textbf{Complex inductions can be specified by grammar.}
\end{quote}
