\begin{quote}
% As my philosophy  colleague Professor Dustin Tucker says, 
\emph{A system studied long enough reveals its paradoxes.}\\
~\hfill-- Dustin Tucker
    % \emph{The power of algebra is that every symbol in an equation is a 
    % variable...}
\end{quote}
Algebra's need operators, and operators as far as we have seen take 
inputs and give outputs.  In other words they are functions.  
As far as simplicity goes the following two example functions win 
on the grounds of requiring nothing, not even a domain or codomain.
\begin{align*}
    I(x) & \defeq x & 
    K_c(x) & \defeq c.
\end{align*}
You may call $I$ the identity function and the $K$ constant functions,
a different one for each $c$.  You can put anything into these functions.
Try some substitutions, I tried $I(3)=3$, $I(\clubsuit)=\clubsuit$.
I found $K_3(2)=3$ and $K_3(\clubsuit)=3$ as well.  I even tried 
$K_{\clubsuit}(2)$ and got $\clubsuit$.  I changed $x$ for $y$, 
$I(y)=y$, and $d$ for $c$, $K_d(x)=d$. Next I substituted $x$ for $c$ and got
\begin{align*}
    K_x(x)=x=I(x).
\end{align*}
Now we have a true problem: a constant function should not equal 
an identity function.  This is the \emph{paradox of the trapped (or captured) variable}.

Paradoxes (para = distinct + dox = opinion) are places where logical reasoning
leads to two seemingly opposing conclusions.  We could just as well call this an
inconsistency and declare the topic dead, but usually mathematics reserves the
word paradox for settings where we could avoid the inconsistency by revisiting
some fundamental notion and choosing a narrower interpretation.  True
inconsistency we reserve for cases where our only option is to throw out the
system.  For example, we could attempt to withhold some options that thus
prevent one of the two opinions to surface.  It is not a philosophically
satisfying resolution, it is religion---a belief. As with most religions there
will be schisms.  I will here follow the church, that is Alanzo Church.
