What is the difference between $x(x+3)^c$ and $f(x)=x(x+3)^c$.  Suppose we use 
the rules to substitute $c\defeq 2$, we get $x(x+3)^2$, $c\defeq \pi$ gives 
use $x(x+3)^{\pi}$.  For $c\defeq d$, we want $x(x+3)^d$.  
Now try $c\defeq x$.  Would you really 
convert this to the function $x(x+3)^x$?  Maybe, but it is more likely that 
we see $x$ and $c$ as distinct, that is, $c$ is constant to $x$.
Something about substitution does not follow induction as cleanly as we have described so far.
If you don't see this as a problem yet lets look at something more basic.

Lets slow down to see what happened.  Set:
\begin{align*}
    I(x) & = x & 
    K_c(x) & = c.
\end{align*}
You may call $I$ the identity function and the $K$ constant functions.
% but with excessive mathematical training you are likely to overthink 
% the situation.  These two formulas have no use for domains and codomains for 
% instance.  We can safely use these even when there are no sets or mathematics 
% around to make sense of deeper concepts.
Try some substitutions, I tried $I(3)=3$, $I(\clubsuit)=\clubsuit$.
I found $K_3(2)=3$ and $K_3(\clubsuit)=3$ as well.  I even tried 
$K_{\clubsuit}(2)$ and got $\clubsuit$.  I changed $x$ for $y$, 
$I(y)=y$, and $d$ for $c$, $K_d(x)=d$. Next I substituted $x$ for $c$ and got
\begin{align*}
    K_x(x)=x=I(x).
\end{align*}
Now we have a true problem: a constant function should not equal 
an identity function.\footnote{Functions in this sense are so primitive 
they have no domains and codomains.  You can put anything into these functions.}
This is the paradox of the trapped variable.

As my philosophy  colleague Professor Dustin Tucker says, 
``A system studied long enough reveals its paradoxes.'' 
Paradoxes (para = distinct + dox = opinion) are inconsistencies that you can avoid by revisiting the scope 
of your definitions.  Just withhold some options and you wont end up with 
two different options.  It is not a philosophically satisfying resolution,
which is why most paradox hacks lead to schisms. 

So what is the root cause of our paradox of the trapped variable?
The answer are bound (local) verses free variables. 
