As a start, it matters first to admit that functions do not need 
domains.  Surprised? Disagree?    Suppose in high school we asked 
a student to  explore
\[
    f(x) = \frac{\sqrt{e^x-1}}{x^2-x-6}.
\]
We might pass this to a graphing calculator, or try some inputs. Along the way
we see the calculator gives up on plotting values $x=(-\infty,0],3$.  Yet, all
we gave it was the formula without prior knowledge of these conditions.  
Be honest with yourself, wouldn't you have thought you could use some of 
those numbers until you tried it as well?  

This problem did not exist in the 1900's.  It is a recent out-growth of strong 
personalities.  Luminaries like David Hilbert urged mathematics to unifty 
around a single ``deductive system''.  Functions no longer did anything, they 
were just reduced to pairing up the before and after photographs.  Like stumbling 
onto a crime scene where you know what was taken but had no clues about the robbers 
or how they pulled it off.  This crime-scene styled view of functions was 
helped along by the homogenizing tendencies of the Bourbaki who recorded the 
first true textbooks an ``all'' mathematics.
Post World War II college enrollment spikes then triggered 
competition for growing textbook markets and monopolies soon emerged crowding 
out less familiar accounts.  Some like me resonate however with the following 
mindful lament.
\begin{quote}
    A function that doesn't function,\\
    should not be called a function.
    \hfill Thorsten Altenkirch
\end{quote}

It is too much of a fight to reclaim the 
name of ``function'' as whatever process you just observed without a domain 
or codomain.  A suitable retreat is to call this an \emph{operator} because 
it operates on data and perhaps changes it, which after all is what a function should have done.

